---
title: "Covid-19 Data"
output:
  pdf_document: default
  html_document: default
  word_document: default
date: "2022-11-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**INTRO**

I will be going through the Covid-19 data. The primary question I am interested in is what was the effect of the Covid-19 vaccine on Covid-19 cases and deaths. I will be going through the Covid-19 data that was looked in class and Vaccine data from the site Listed below.

My hypothesis is that the number of deaths_per_thou should be lower in the states with the highest vaccination rates.

In order to build this RMD file we the following packages: library(tidyverse) library(lubridate)

NOTE:  I LEFT ALL OF THE HELPER/ERROR MESSAGES ON, AS I THOUGHT IT WOULD BE HELPFUL.  

# Tidy the Data

```{r TestCovidData, include=TRUE}
    library(tidyverse)
    library(lubridate)

    # Just assign a name to a variable. (str)
    url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

    # Same as above... create an array with 4 names.
    file_names <- c("time_series_covid19_confirmed_US.csv", "time_series_covid19_deaths_US.csv" )

    # concatinate each one with the url_in.  
    urls <- str_c(url_in,file_names)
    
    urlvaccine <- "https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/vaccinations/us_state_vaccinations.csv"
    # read those and assign to variables.
    
    us_cases <- read_csv(urls[1])
    us_deaths <- read_csv(urls[2])
    Vaccine_data <- read_csv(urlvaccine)

    uid <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

    popul <- read_csv(uid)
    
    # Now we are going to join the population in global.

    # DO THE SAME THINGS WITH US CASES:
    us_cases <- us_cases %>% pivot_longer(cols = -(UID:Combined_Key), names_to = "date", values_to = "cases")
    us_cases <- us_cases %>% select(Admin2:cases)
    us_cases <- us_cases %>% mutate(date = mdy(date))

    # DO THE SAME WITH US DEATHS (INCLUDING POPULATION)
    
    us_deaths <- us_deaths %>% pivot_longer(cols = -(UID:Population), names_to = "date", values_to = "deaths")
    us_deaths <- us_deaths %>% select(Admin2:deaths)
    us_deaths <- us_deaths %>% mutate(date = mdy(date))
    us_deaths <- us_deaths %>% select(-c(Lat,Long_))
    
    # Combine the two
    US_by_state <- us_cases %>% full_join(us_deaths)
```

# Transform

Lets add a few more columns and start to look at some things of interest.

```{r TestCovidData2, include=TRUE}
    # Started another one to now go through and visualize and analize some of the data above

    # Group the data by the following:  
    US_by_state <- US_by_state %>% group_by(Province_State, Country_Region, date)
    
    # add the following:
    US_by_state <- US_by_state %>% summarise(cases = sum(cases), deaths = sum(deaths), Population = sum(Population))
    
    # Add another column
    US_by_state <- US_by_state %>% mutate(deaths_per_mill = deaths *1000000 / Population) 
    US_by_state <- US_by_state %>% mutate(cases_per_mill = cases *1000000 / Population) 
    # Ungroup
    US_by_state <- US_by_state %>% ungroup()

    US_state_totals <- US_by_state %>% group_by(Province_State) %>% summarise(deaths=max(deaths), cases=max(cases), population = max(Population), cases_per_thou=1000*cases/ population, deaths_per_thou=1000* deaths / population) %>% filter(cases>0, population>0)
    
    # Lets add the Vaccine Data to the US_state_totals
    
    VaccineTotals <- Vaccine_data %>% select(date,location, people_fully_vaccinated)
    
    # Lets get the totals for each state
    
    VaccineTotals <- VaccineTotals %>% group_by(location) %>% summarise(People_fully_vaccinated=max(people_fully_vaccinated, na.rm=TRUE))

    # rename location to Province_State
    VaccineTotals <- VaccineTotals %>% rename(Province_State = location)
    
    # Lets combine this with the US_State_totals
    
    US_state_totals <- US_state_totals  %>% full_join(VaccineTotals)
    
    # Lets add another column for percent 
    
    US_state_totals <- US_state_totals  %>% mutate(VaccinePercent = People_fully_vaccinated/population ) 

    # Lets Cleanup Vaccine data as well.  
    
    Vaccine_data <- Vaccine_data %>% select(date, location, people_fully_vaccinated )
    Vaccine_data <- Vaccine_data %>% rename(Province_State = `location`)

```

# Interesting Charts

Now lets take a look at some interesting states.

```{r totals1, include=TRUE}
    # Here is a list of states with the most deaths.

    US_state_totals %>% slice_max(deaths_per_thou, n=10)

```

Let us take a look at some states with the lowest death rates:

```{r totals2, include=TRUE}

    US_state_totals %>% slice_min(deaths_per_thou, n=10)
```

Lets take a look at the Vaccination rates

```{r totals3, include=TRUE}

    US_state_totals %>% slice_max(VaccinePercent, n=10)

```

Lets Take a look at the states with the minimum Vaccine Rates

```{r totals4, include=TRUE}

    US_state_totals %>% slice_min(VaccinePercent, n=10)

```

# ANALYSIS

## From the above date lets take a look at the following states:##

I picked Massachusetts as it was a state with a pretty large population.. compared to the rest of the states.

Massachusetts 24249 2220962 6892503 322.2287 3.5181704 5810671

## I thought it was important to pick something that was quite a bit further away from the Northeast. ##

Hawaii 1834 380098 1415872 268.4551 1.2953148 1154399

## I picked these as these were the two states with the lowest vaccination rates. They also had population over one Million. ##

Alabama 21001 1642062 4903185 334.8970 4.283134 2607186 
Missouri 22825 1773682 6626371 267.6702

```{r StateGraphs, include=TRUE}
    
# Lets look at the us totals as well.  This was done in class
    
    us_totals <- US_by_state %>% group_by(Country_Region, date) %>% summarize(cases=sum(cases), deaths=sum(deaths), Population = sum(Population)) %>% mutate(deaths_per_mill = deaths * 1000000/Population)

    # Ungroup them. 
    us_totals <- us_totals %>% ungroup()

    # SHOW THE GRAPH OF DEATHS VS. CASES
    
    us_totals %>% filter(cases> 0) %>% ggplot(aes(x=date, y=cases)) +geom_line(aes(color = "cases")) +geom_point(aes(color = "cases")) + geom_line(aes(y=deaths, color = "deaths")) +geom_line(aes(color = "deaths")) +geom_point(aes(color = "deaths")) + scale_y_log10() + theme(legend.position="bottom", axis.text.x = element_text(angle=90)) +labs(title = "Covid19 in US", Y = NULL)

```

## Visualize

It looks like overall the cases and death rates are flat. This was what we would expect from looking at the US Data above.

```{r StateGraphs1, include=TRUE}

# Lets take a look at the 4 states we picked above.  Two from the lowest catagory and two from the highest catagory of vaccinations.

    US_by_state %>% filter(cases > 0, deaths > 0, Province_State == "Alabama") %>% ggplot(aes(x=date, y=cases)) +geom_line(aes(color="cases")) +geom_line(aes(y=deaths, color="deaths")) + scale_y_log10() +labs(title = "Covid19 in Alabama", Y = NULL)
    
        US_by_state %>% filter(cases > 0, deaths > 0, Province_State == "Missouri") %>% ggplot(aes(x=date, y=cases)) +geom_line(aes(color="cases")) +geom_line(aes(y=deaths_per_mill, color="deaths_per_mill")) + scale_y_log10() +labs(title = "Covid19 in Missouri", Y = NULL)
        
        US_by_state %>% filter(cases > 0, deaths > 0, Province_State == "Massachusetts") %>% ggplot(aes(x=date, y=cases)) +geom_line(aes(color="cases")) +geom_line(aes(y=deaths_per_mill, color="deaths_per_mill")) + scale_y_log10() +labs(title = "Covid19 in Massachusetts", Y = NULL)

    US_by_state %>% filter(cases > 0, deaths > 0, Province_State == "Hawaii", date > "2022-01-01") %>% ggplot(aes(x=date, y=cases)) +geom_line(aes(color="cases")) +geom_line(aes(y=deaths_per_mill, color="deaths_per_mill")) + scale_y_log10() +labs(title = "Covid19 in Hawaii", Y = NULL)

    
 
```
# VISUALIZE

    Now lets look at the Vaccine data and new Deaths and New cases.  
``` {r MyAnalysis, include=TRUE}


# Since all we care about is people who have been fully_vaccinated.

    #Vaccine_data <- Vaccine_data %>% select(date, location, people_fully_vaccinated )
    #Vaccine_data <- Vaccine_data %>% rename(Province_State = `location`)

# Now lets go ahead and combine this with the states total.  

    US_by_state_vaccine <- US_by_state %>% full_join(Vaccine_data,by=c("Province_State","date"))

    # Now lets just look at the new cases
    US_by_state_vaccine <- US_by_state_vaccine %>% mutate(new_cases = cases - lag(cases), new_deaths=deaths - lag(deaths), new_vaccine = people_fully_vaccinated - lag(people_fully_vaccinated) )

    US_by_state_vaccine <- US_by_state_vaccine %>% mutate(Percent_deaths = new_deaths/new_cases *100)

    US_by_state_vaccine <- US_by_state_vaccine %>% mutate(vaccine_per_hundred = 1000 * people_fully_vaccinated/Population)

    US_by_state_vaccine <- US_by_state_vaccine %>% mutate(cases_per_hundred = 1000 * cases/Population)

    US_by_state_vaccine %>% filter(new_deaths > 0, people_fully_vaccinated > 0, new_cases > 0, Province_State == "Alabama") %>% ggplot(aes(x=date, y=new_cases)) +geom_line(aes(color="new_cases")) +geom_line(aes(y=new_deaths, color="new_deaths"))+ggtitle("New Cases vs. Deaths in Alabama")

    US_by_state_vaccine %>% filter(new_deaths > 0, people_fully_vaccinated > 0, new_cases > 0, Province_State == "Missouri") %>% ggplot(aes(x=date, y=new_cases)) +geom_line(aes(color="new_cases")) +geom_line(aes(y=new_deaths, color="new_deaths"))+ ggtitle("New Cases vs. Deaths in Missouri")

    US_by_state_vaccine %>% filter(new_deaths > 0, people_fully_vaccinated > 0, new_cases > 0, Province_State == "Massachusetts") %>% ggplot(aes(x=date, y=new_cases)) +geom_line(aes(color="new_cases")) +geom_line(aes(y=new_deaths, color="new_deaths"))+ ggtitle("New Cases vs. Deaths in Massachusetts")

    US_by_state_vaccine %>% filter(new_deaths > 0, people_fully_vaccinated > 0, new_cases > 0, Province_State == "Hawaii") %>% ggplot(aes(x=date, y=new_cases)) +geom_line(aes(color="new_cases")) +geom_line(aes(y=new_deaths, color="new_deaths")) + ggtitle("New Cases vs. Deaths in Hawaii")

       
    ```

# MORE VISUALS

Lets see if the percent of deaths vs. new cases in each of the above states.

So our hypothesis was that states with the greater number of vaccinated residents should have lesser number of residents dying. So we are going to take a look at each of the states and see if that's true.

```{r VaccineS, include=TRUE}

# Some things to note.  I took a look at all numbers after 2021-07-01.  Mostly because we started vaccinations in January of 2021 and it takes time to Vaccinate people.

# Lets take a look at Alabama First.  
US_by_state_vaccine %>% filter(new_deaths > 0, people_fully_vaccinated > 0, new_cases > 0, Province_State == "Alabama", Percent_deaths < 25, date > '2021-07-01') %>% ggplot(aes(x=date, y=Percent_deaths)) +geom_line(aes(color="Percent_deaths"))+labs(title = "Percent of People dying vs. Cases in Alabama", Y = NULL)

# Lets take a look at Missouri.  

US_by_state_vaccine %>% filter(new_deaths > 0, people_fully_vaccinated > 0, new_cases > 0, Province_State == "Missouri", Percent_deaths < 25, date > '2021-07-01') %>% ggplot(aes(x=date, y=Percent_deaths)) +geom_line(aes(color="Percent_deaths"))+labs(title = "Percent of People dying vs. Cases in Missouri", Y = NULL)



US_by_state_vaccine %>% filter(new_deaths > 0, people_fully_vaccinated > 0, new_cases > 0, Province_State == "Massachusetts", Percent_deaths < 25, date > '2021-07-01') %>% ggplot(aes(x=date, y=Percent_deaths)) +geom_line(aes(color="Percent_deaths"))+geom_line(aes(color="Percent_deaths"))+labs(title = "Percent of People dying vs. Cases in Massachusetts", Y = NULL)

US_by_state_vaccine %>% filter(new_deaths > 0, people_fully_vaccinated > 0, new_cases > 0, Province_State == "Hawaii", Percent_deaths < 25, date > '2021-07-01') %>% ggplot(aes(x=date, y=Percent_deaths)) +geom_line(aes(color="Percent_deaths"))+geom_line(aes(color="Percent_deaths"))+labs(title = "Percent of People dying vs. Cases in Hawaii", Y = NULL)

```
# ANALYSIS

As we can clearly see the death rate in states with high vaccination is quite a bit.  Lets just take a look at Alabama and Massachusetts and you can see the chances of you dying from Covid-19 goes down dramatically.  

## BIAS

Personal Bias I think I picked Massachusetts as I have ties to that state. 

Now sure how each state are collecting there data.

Demographics can also play a big role. 

